---
title: "HW2-EDA with R - Part1: Plotting"
author: "Michael Kazmierski, Erin Barr"
date: "January 22, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.
You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset.
The last file, **county_MI.csv**, is a file that I created and just contains
data from Michigan. The R script, **census_data_prep.R** was used to create
the file as well as to merge information from the **us_census_bureau_regions_and_divisions.csv** file with the county data.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class,
you should use a combination of markdown text (be concise, no need to write tons
of text) to explain what you are doing and R code chunks to actually do it. When
you are done, use the "Knit" button to generate an HTML file from your R
Markdown. You'll be submitting BOTH the completed R Markdown file as well as the
generated HTML file. You'll notice that I've left some empty code chunks as well
as some partially filled in code chunks. Of course, feel free to add more code
chunks as needed.

### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}

county_df <- read.csv("./data/county.csv") 
county_MI <- read.csv("./data/county_MI.csv")
```

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. Are there any NA values? What 
differences do you notice between the structure of the two data frames.

```{r}
summary(county_MI)
```

```{r}
summary(county_df)

```
There are 3 NAs in this data frame. 
county_MI has 41 variables while the data frame, county_df, above has 40.

The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.

### Problem 2 - Basic plots for Income

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 

Write a short paragraph describing the distribution of median incomes.

```{r}
ggplot(county_df, aes(county_df$Income)) + geom_histogram(fill="violet", alpha=0.5) + labs(x="Income")
ggplot(county_df, aes("MedianIncome",county_df$Income)) + geom_boxplot(color="#B59A57") + labs(y = "Dollars", x="")
ggplot(county_df, aes(x=county_df$Income)) + geom_density(color="#3d6e27")
ggplot(county_df, aes(x="Income", y=county_df$Income)) + geom_violin(trim=FALSE, fill="grey80", color="#3366ff") + labs(y="Dollars", x="")
```
The sample below is somewhat close to a normal distribution suggesting that the sample is overwhelming middle class. However, there is a slight positive skewness. This indicates a larger gap between the wealthy and middle class than there is between the poor and the middle class. 
### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US.

```{r histo_totpop}
ggplot(county_df, aes(county_df$TotalPop)) + geom_histogram(fill="light blue", alpha=0.7) + labs(x="Total Population")
```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
ggplot(county_df, aes(county_df$TotalPop)) + geom_histogram(fill="light blue", alpha=0.7) + labs(x="Total Population", y="Number of Counties", title="Distribution of Population for US Counties" )
```

**Hacker Extra** - Also create a kernel density plot overlaid on a histogram of
Income. 

Repeat the histogram but only for counties with less than 2.5 million people.

```{r histo_lt2.5m}
county_df_2.5 <- county_df %>%
  filter(TotalPop < 2500000)

ggplot(county_df, aes(county_df$TotalPop)) +
  geom_density(aes(y=250000*..count..)) +
  geom_histogram(fill="light blue", alpha=0.9, binwidth = 250000) + 
    labs(x="Total Population", y="Number of Counties", title="Distribution of Population for US Counties" )
  
```

```{r histo_lt2.5m}
ggplot(county_df_2.5, aes(county_df_2.5$TotalPop)) + geom_histogram(fill="light green", alpha=0.9) + labs(x="Total Population", y="Number of Counties", title="Distribution of Population for US Counties with Less than 2.5M" )
```


### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped box plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?


```{r bp_inc_by_peninsula}

peninsula=rep(c("Lower","Upper"))

ggplot(county_MI, aes(y=county_MI$Income, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
  labs(x="Peninsula", y="Income", title="Income by Peninsula" )

```
Income differs between the two peninsulas.  The lower peninsula has a higher median income with greater variablility than the upper peninsula.

Create several more boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r more_boxplots}
ggplot(county_MI, aes(y=county_MI$TotalPop, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
    guides(color=FALSE)+
  labs(x="Peninsula", y="Population", title="County Population by Peninsula" )

ggplot(county_MI, aes(y=county_MI$Hispanic, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
    guides(color=FALSE)+
  labs(x="Peninsula", y="Hispanic Population", title="Hispanic Population by Peninsula" )

ggplot(county_MI, aes(y=county_MI$White, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
    guides(color=FALSE)+
  labs(x="Peninsula", y="White Population", title="White Population by Peninsula" )

ggplot(county_MI, aes(y=county_MI$Black, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
    guides(color=FALSE)+
  labs(x="Peninsula", y="Black Population", title="Black Population by Peninsula" )

ggplot(county_MI, aes(y=county_MI$Poverty, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
    guides(color=FALSE)+
  labs(x="Peninsula", y="Poverty", title="Poverty by Peninsula" )

ggplot(county_MI, aes(y=county_MI$MeanCommute, x=county_MI$peninsula, fill=peninsula)) + 
    geom_boxplot()+
    guides(color=FALSE)+
  labs(x="Peninsula", y="Mean Commute", title="Mean Commute by Peninsula" )

```
Median county population is greater in the lower peninsula than the upper peninsula, with significantly higher variation.  Similarly, Hispanic Populations in each county is higher in the loweer peninsula than the upper peninsula.  However, the median white population by county is lower in the lower peninsula than the upper, though with a much higher variance - there are outlier counties that have a relatively low white population in the lower peninsula.  We see this information reflected in Black Population by county in each peninsula, with outlier counties contributing to a much higher variability in the lower peninsula.  

Comparing the two peninsulas on poverty there is a much greater variability in the lower peninsula than the upper peninsula.  However, when looking at the median number of people in poverty in the lower peninsula is slightly higher.  

Commuters in the lower peninsula have a longer commute than the commuters in the upper peninsula.  The median of the lower peninsula is higher than the majority of the lower peninsula's commutes.

### Problem 5 - How do regions of the country compare?

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how some variable of interest to you compares by region. For 
example, you could explore how the regions compare with respect to the
distribution of MeanCommute time. Make sure your graph has a title and good axis
labels.

```{r faceted_histo}
ggplot(county_df, aes(county_df$MeanCommute)) + 
  geom_histogram(fill="light green", alpha=0.9) + 
  labs(x="Mean Commute", y="Number of Counties", title="Mean Commute by County" )+
  facet_grid(Region~ .)
```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Region in the US. Make sure
your graph has a title and good axis labels.

```{r bar_counties_by_region}
Midwest <- (sum(county_df$Region == "Midwest"))
South <- (sum(county_df$Region == "South"))
Northeast <- (sum(county_df$Region == "Northeast"))
West <- (sum(county_df$Region == "West"))

Region <- c("Midwest", "South", "Northeast", "West")
CountyCount <- c(Midwest, South, Northeast, West)

Regions <- data.frame(Region, CountyCount)

ggplot(Regions, aes(Region, CountyCount)) + 
  geom_col(stat=Region, fill="orange", alpha=0.7) + 
  labs(x="Region", y="Number of Counties", title="Number of Counties by Region" )+
  coord_flip()

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}
ggplot(county_MI, aes(county_MI$MeanCommute, county_MI$Income)) +
  geom_point(color = "blue") +
  labs(x="Mean Commute", y="Income", title="Income vs Mean Commute in Michigan Counties")


```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula.

```{r xy_inc_commute_peninsula}
ggplot(county_MI, aes(county_MI$MeanCommute, county_MI$Income, color=peninsula)) +
  geom_point() +
  labs(x="Mean Commute", y="Income", title="Income vs Mean Commute in Michigan Counties")
```

Your thoughts on the relationship between these variables?

People in the lower peninsula drive farther and make more while the people in the upper peninsula drive less and make less on average. 



Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

```{r xy_faceted_peninsula}
ggplot(county_MI, aes(county_MI$WorkAtHome, county_MI$Income, color=peninsula)) +
  geom_point() +
  labs(x="Working At Home", y="Income", title="Income vs Working From Home in Michigan Counties")

ggplot(county_MI, aes(county_MI$MeanCommute, county_MI$Carpool, color=peninsula)) +
  geom_point() +
  labs(x="Mean Commute", y="Car Pool", title="Car Pool vs Mean Commute in Michigan Counties")

ggplot(county_MI, aes(county_MI$Unemployment, county_MI$MeanCommute, color=peninsula)) +
  geom_point() +
  labs(x="County Population", y="Unemployment", title="Total Population vs Unemployment in Michigan Counties")
```
  When it comes to working at home versus income the lower peninsula has a wide range of outcomes. Whereas te upper peninsula only semes to have the lower end of the specturm of the income scale. This shows that if you want to work from home and still make a higher income, you may want to live in the lower peninsula. 
  Commuting the upper peninsula tends to stay the under 22 minutes, which means they have a shorter drive to work. While the lower peninsula doesn't seem to car pool much unless they have a 20 minute commute or longer. The majority of the lower peninsula mean commute seems to be between 20-30 minutes while the lower peninsula is 17-22 minutes. While the number of car poolers tends to be about the same for both peninsulas. 
  Unemployment for the county population for both peninsulas have the same tendency when it comes to unemployment. The only differnce is that the lower peninsula has a higher population therefore there is more occurance of it and so their dots on the graph are above the dots of the upper peninsula. 

Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}
ggplot(county_df, aes(county_df$TotalPop, county_df$Income)) +
  geom_point(color = "purple") +
  labs(x="Total Population", y="Income", title="Income vs Total Population")
```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.

Outliers, high skewness, and high variation in the data can make it difficult to interpret or discover patterns.  Transforming data using the logarithmic method reduces variance and sets the data in a normal distribution.  Additionally, transforming data may improve the regression analysis itself, such as the error distribution.

```{r}
county_df$TotalPop_log10 <- log10(county_df$TotalPop)
county_df$Income_log10 <- log10(county_df$Income)
```

```{r}
ggplot(county_df, aes(county_df$TotalPop_log10, county_df$Income_log10)) +
  geom_point(color = "dark green") +
  labs(x="Total Population (log10)", y="Income (log10)", title="Income vs Total Population")
```

```{r}
ggplot(county_df, aes(county_df$Income_log10)) + geom_histogram(fill="black", alpha=0.5) + labs(x="Income (log10)")
#Demonstrates normal distribution of Income after log10 transformation.
```



### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.

```{r}
county_df$WhitePercent <- (countydf)
```

